---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# datadist

<!-- badges: start -->
<!-- badges: end -->

The goal of datadist is to ...

## Installation

You can install the development version of datadist from [GitHub](https://github.com/) with:

``` {r}
# install.packages("pak")
pak::pak("nlharms414/datadist")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(datadist)
## basic example code
```

### Diamonds Data from kaggle

The object `kgl_diamonds` contains five data of the 320 data sets available on kaggle when searching the platform for `diamonds` (Feb 2026).

At a first glance, the dimensions of these datasets are similar to each other, but not identical.

```{r}
data("kgl_diamonds")
kgl_diamonds |> purrr::map(.f = dim)
```


The dimensions are also very similar to the `diamonds` dataset in `ggplot2`:

```{r}
library(ggplot2)
library(dplyr)
dim(diamonds)
```

Are some of these datasets identical to the `diamonds` data?

```{r}
kgl_diamonds |> purrr::map(.f = function(x) identical(x,diamonds))
```
When we look a bit closer, though, we find that "vittoriogiatti" has strong similarities to the `diamonds` data in ggplot2, as all of the names of the variables coincide and the numeric values are the same, even in the same order.

```{r warning=FALSE}
summary(kgl_diamonds[[3]] - diamonds)
```
This is the same picture we get when we calculate the marginal distances between all pairs of variables between the two datasets:

```{r}
wasserstein <- numdist(kgl_diamonds[[3]], diamonds)
round(wasserstein, 2)
```

We see 0s on the diagonal, indicating that these pairs of variables have the same marginal distribution, i.e. these variables have identical data.
We also see that the difference between `x` and `y` is very small, indicating a high positive correlation between these two measurements, i.e. most diamonds are highly symmetric.


data error: y = 58.9
no diamonds priced between \$1460 and \$1540

```{r}
diamonds |> filter(between(price, 1000, 3000)) |> ggplot(aes(x = price)) + geom_histogram(binwidth=10)
```

This matrix can be summarized in a score calculated as the minimum of the sum of all possible variable to variable mappings:

```{r}
dscore(wasserstein)$score
```
A value of 0 in this `dscore` indicates that all variables in the datasets have identical marginal distributions.


Datasets with an additional variable have the row numbers saved in a first column: 

```{r}
wasserstein_1 <- numdist(kgl_diamonds[[1]], diamonds)
round(wasserstein_1, 2)
```

They are, however, identical to each other:

```{r}
wasserstein_12 <- numdist(kgl_diamonds[[1]], kgl_diamonds[[2]])
round(wasserstein_12, 2)
```

More interestingly, the number of observations in `kgl_diamonds` data sets 4 and 5 are different from the original. 

```{r}
wasserstein_4 <- numdist(kgl_diamonds[[4]], diamonds)
wasserstein_5 <- numdist(kgl_diamonds[[5]], diamonds)
```

Are these maybe submissions that used the script to update the data?

```{r}
kgl_diamonds[[4]] |> filter(between(price, 1400, 1600)) |> ggplot(aes(x = price)) + geom_histogram(binwidth=10)

kgl_diamonds[[5]] |> filter(between(price, 1400, 1600)) |> ggplot(aes(x = price)) + geom_histogram(binwidth=10)

```

Alas, no.

```{r}
kgl_diamonds[[5]] |> ggplot(aes(x = x, y = y)) + geom_point()
```

# Comparison of character/categorical variables

Different situations: comparison of numeric to categorical - not sure how, later

categorical  to categorical (lowish number of categories):

```{r}
full_x <- kgl_diamonds[[5]]$cut |> as.factor() |> ecdf()

table_x <- kgl_diamonds[[5]]$cut |> table() |> ecdf()
# gives different ecdfs because of different orderings


wassersteinXY(kgl_diamonds[[5]], diamonds)


# option 0: count all non-matching strings (for some string distance) (needs the same number of observations)
char_dist <- sum(kgl_diamonds[[1]]$cut != sample(diamonds$cut))/length(kgl_diamonds[[1]]$cut)

# option 1: use dummy encoding (0/1 data.frames) then use datadist on those data frames
dfA <- data.frame(model.matrix(~cut-1, data = kgl_diamonds[[5]]))
dfB <- data.frame(model.matrix(~cut-1, data = diamonds))

dfA$name <- "A"
dfB$name <- "A"



```

Idea for categorical distances:

1. check if X is equal to Y (only works for the same number of observations)
2. use dummy encoding for X and Y and run numscore


