---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# datadist

<!-- badges: start -->
<!-- badges: end -->

The goal of datadist is to ...

## Installation

You can install the development version of datadist from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("nlharms414/datadist")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(datadist)
## basic example code
```

### Diamonds Data from kaggle

The object `kgl_diamonds` contains five data of the 320 data sets available on kaggle when searching the platform for `diamonds` (Feb 2026).

At a first glance, the dimensions of these datasets are similar to each other, but not identical.

```{r}
data("kgl_diamonds")
kgl_diamonds |> purrr::map(.f = dim)
```


The dimensions are also very similar to the `diamonds` dataset in `ggplot2`:

```{r}
library(ggplot2)
dim(diamonds)
```

Are some of these datasets identical to the `diamonds` data?

```{r}
kgl_diamonds |> purrr::map(.f = function(x) identical(x,diamonds))
```
When we look a bit closer, though, we find that "vittoriogiatti" has strong similarities to the `diamonds` data in ggplot2, as all of the names of the variables coincide and the numeric values are the same, even in the same order.

```{r warning=FALSE}
summary(kgl_diamonds[[3]] - diamonds)
```
This is the same picture we get when we calculate the marginal distances between all pairs of variables between the two datasets:

```{r}
wasserstein <- datadist(kgl_diamonds[[3]], diamonds)$dist
round(wasserstein, 2)
```
Datasets with an additional variable have the row numbers saved in a first column: 

```{r}
wasserstein_1 <- datadist(kgl_diamonds[[1]], diamonds)$dist
round(wasserstein_1, 2)
```

They are, however, identical to each other:

```{r}
wasserstein_12 <- datadist(kgl_diamonds[[1]], kgl_diamonds[[2]])$dist
round(wasserstein_12, 2)
```
