---
title: "Whodunit?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Whodunit?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette features a "whodunit" style problem to demonstrate the `datadist` package use. 

# Palmer Penguins mix up

Uh oh! The original Palmer Penguins (`penguins`) data set has been badly misshapen. Variable names are missing, categories for categorical variables have been replaced with letters, numerical variables have been scaled, and columns have been reordered. It is our job to use functions from the `datadist` package to determine what what done to the original `penguins` data set that resulted in such a catastrophe.

First, we'll want to install the package.

```{r}
library(datadist)
```

## Using `chardist`, `dscore`, and `chardist_XY`

Let's first figure out what happened with the categorical variables in the data set. First, we can use `chardist` to generate a distance matrix for every pairwise combination of categorical variables between the two data sets. 

```{r}
peng_dist <- chardist(penguins,penguins_whodunit)
peng_dist
```

This result is already pretty telling, but lets use `dscore` to look at variable mappings just to confirm things. The `dist` column will give us a distance score between the two variables from each data set that are the best match.

```{r}
dscore(peng_dist)$mappings
# dscore() actually returns a list of items
# use $ + tab to see what else it returns!
```

Great! Now we know which variables are which. But, recall that all of the categorical variables in the data set had been recoded with letters. Luckily, `chardist_XY` can help us solve that problem. We can use it on the variable pairs above to tell us which categories from the original data set are represented by each letter. `chardist_XY` is designed to work on single variable comparisons, and the results returned are a combination of `chardist` and `dscore`.

```{r}
chardist_XY(penguins$species,penguins_whodunit$Var4)$distmat$mappings
chardist_XY(penguins$sex,penguins_whodunit$Var5)$distmat$mappings
chardist_XY(penguins$island,penguins_whodunit$Var8)$distmat$mappings
```

So, how do we interpret the results in their entirety? Well, the use of `chardist` and `dscore` told us that `Var4` in `penguins_whodunit` is actually `species` from the original `penguins` data set, for example (and so on and so forth for the other mappings). Then, if we take a closer look at the actual factors in `Var4` and `species`, we can find that "A" was actually "Adelie", "B" was actually "Chinstrap", and "C" was actually "Gentoo". We can look more closely at the other variable mappings the same way. Now that we've got the categorical variables covered, let's move on to the numerical ones.

## Using `numdist` and `dscore`

With the numerical variables, the goal here is to simply match the variables. Let's run `numdist` on the two data sets and see what happens.

```{r}
numdist(penguins,penguins_whodunit)
```

Huh. That's weird- none of the distance values indicate that that any of the numerical variables in `penguins_whodunit` are good matches for the variables in the original `penguins` data set. What's going on here?

Looking at the numerical variables in the `penguins_whodunit` data set, we can tell they have all been normalized. So, when we use `numdist` to look at a distance matrix of all the numerical variables, we want to set the `scale` option to `TRUE`. Why? `numdist` relies on `wasserstein_XY`, a function that computes the Wasserstein distance between two variables. Since the computation of the Wasserstein distance relies on the distance between the actual data values themselves, we should scale the variables in both data sets before computing the Wasserstein distance to get them both on the same scale. Then, the distance calculation between the two variables will be more accurate. 

```{r}
numdist(penguins,penguins_whodunit, scale = T)
```

Ah, that looks much better. Though, with all the decimals it's not super easy to tell which variables match to which just by looking at it. This is where applying `dscore` to look at the mappings will come in handy.

```{r}
dscore(numdist(penguins,penguins_whodunit, scale = T))$mappings
```

`dscore` determined that the mappings seen above result in the lowest over all distance score for the numerical variables. Thus, we can interpret the results similarly to what we did with the categorical variables. Here, `Var1` in the `penguins_whodunit` data set was originally the `bill_len` variable in the original data set, and likewise we can interpret all the other variable mappings.

It's noted here that none of the scores between the best matched variables are *exactly* zero, though they are close. This again has to do with the scaling; the variables in `penguins_whodunit` were scaled to begin with, and because we used the `scale=T` argument when we ran `numdist`, those variables were scaled *again*. So, while scaling the variable a second time doesn't change the values by much and allows for proper comparison, it is just enough change to result in the non-zero scores we see above.

## So, whodunit??

That concludes our sleuthing! By using a combination of `datadist` functions, we have now gathered the information needed to figure out which variables were which, and subsequently, we have the information we need to convert `penguins_whodunit` back to the original `penguins` data set. And hopefully along the way, you've learned a little more about the `datadist` functions and how to use them.




